\texttt{thornado}-transport: IMEX schemes for two-moment neutrino transport respecting Fermi-Dirac statistics

### 2018-11-07 Documents/M1/Astronum_2018/sections/SpatialDiscretization.tex
Changed constraint-preserving to realizability-preserving.

In this section we discretize the two-moment model with a simple first-order finite volume method to illustrate how the closure affects the realizability-preserving property of the scheme.  
By assuming that the moments at time level $t^{n}$ ($\bcM^{n}$) satisfy the bounds in Eq.~\eqref{eq:MomentsBounds}, our goal is to identify sufficient conditions such that the moments at time level $t^{n+1}$ ($\bcM^{n+1}$) also satisfy the bounds.  
To simply the illustration, we limit the discussion to one spatial dimension and employ a uniform Cartesian mesh.  
(The extension to multiple spatial dimensions and high-order discretization using the discontinuous Galerkin method is given in \cite{chu_etal_2018})

>>>>

In this section we discretize the two-moment model with a simple first-order finite volume method to illustrate how the closure affects the realizability-preserving property of the method.  
By assuming that the moments at time level $t^{n}$ ($\bcM^{n}$) satisfy the bounds in Eq.~\eqref{eq:MomentsBounds}, our goal is to identify sufficient conditions such that the moments at time level $t^{n+1}$ ($\bcM^{n+1}$) also satisfy the bounds.  
To simply the illustration, we limit the discussion to one spatial dimension and employ a uniform Cartesian mesh.  
(The extension to multiple spatial dimensions and high-order discretization using the discontinuous Galerkin method is given in \cite{chu_etal_2018})

### 2018-11-07 Documents/M1/Astronum_2018/sections/TimeIntegration.tex
Changed constraint-preserving to realizability-preserving.

Since the set of realizable moments is convex, convex-invariant schemes, which maintain states in a convex set, can be used to design realizability-preserving schemes for the two-moment model. 
Ideally, the scheme should also be high-order accurate and work well in the asymptotic diffusion limit (characterized by frequent collisions and long time scales).  
The following discussion considers the construction of such convex-invariant schemes.

@@

\subsection{Constraint-Preserving Implicit-Explicit Schemes}

@@

By requiring the IMEX scheme be GSA, $\vect{u}^{n+1} = \vect{u}^{(s)}$.
Then using those Lemma proved in \cite{chu_etal_2018}, we can find those constraints, such as Eq.~\eqref{eq:convexInvariant} for type~ARS, that $c_{ij}$ and $\hat{c}_{ij}$ (or $a_{ij}$ and $\tilde{a}_{ij}$) need to satisfy to ensure $\vect{u}^{(i)}$, Eq.~\eqref{eq:imexStagesRewrite}, be realizable.
Hence $\vect{u}^{(i)}$ and $\vect{u}^{n+1}$ are realizable and the IMEX is constraint-preserving.

@@

Unfortunately, no coefficients can be found for standard IMEX schemes under the constraint-preserving constraints and the accuracy constraints for 2nd-order or higher. 
To circumvent this problem, some correction steps are introduced after the assembly step, Eq.~\eqref{imexIntermediate}, such as Hu et al.\cite{hu_etal_2018}.
However, the correction steps impose extra constraints for realizability on time step.
In the circumstance that we have, correction steps can ruin the efficiency we gain from the IMEX scheme.
To keep things simple, we aim for 1st-order in general, high-order in streaming limit, diffusion accurate, constraint-preserving IMEX schemes.

\subsection{Diffusion Accurate, Constraint-Preserving Implicit-Explicit Schemes}

Diffusion accurate is an important property the time integration should have.
In the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$.
The moment behavior are governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
In the context of IMEX schemes, the above relationships imply that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold. 
That's
\begin{equation}
   \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s,
\end{equation}
where $\vect{e}_{i}$ is the $i$th column of the $s\times s$ identity matrix.
\rc{It can be rewrite to $c=\tilde{c}$.}
We have proved in \cite{chu_etal_2018} that only type~ARS IMEX schemes can be diffusion accurate and constraint-preserving at the same time.

@@

Combining the conditions we have, the 1st-order in general, high-order in streaming limit, diffusion accurate, constraint-preserving IMEX schemes we need are type~ARS IMEX schemes satisfying the following constraints:
\begin{enumerate}
    \item Consistency of the implicit coefficients:
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item High-order accuracy in the streaming limit, 2nd-order for example:
    \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2},
      \label{eq:orderConditionsEx}
    \end{equation}
    \item Diffusion accurate: \rc{$c=\tilde{c}$.}
    \begin{equation}
      \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Convex-invariant:
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
      \label{eq:convexInvariant}
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.
    \item Less than five stages ($s\le4$) \label{cod:statges}.
    \item Globally stiffly accurate: $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$. 
\end{enumerate}  
The \ref{cod:statges} constraint is given by efficiency consideration.
We call the IMEX scheme satisfying the above conditions {PD-ARS}. (Definition 3 in \cite{chu_etal_2018}.)
We give two optimized PD-ARS schemes here: PD-ARS with SSPRK2 and  PD-ARS with SSPRK3. 
They have second- and third-order accuracy in the streaming limit, respectively.
\subsubsection{PD-ARS with SSPRK2}
Here we give the optimized PD-ARS with SSPRK2 in the standard double Butcher tableau form: explicit tableau on the left ($\tilde{A}$), implicit tableau on the right ($A$):
\begin{equation}
  \begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2 & 1/2 \\ \hline
  	  & 0 & 1/2 & 1/2
  \end{array}
\end{equation}
For this scheme, $c_{\mbox{\tiny Sch}}= 1$ and only two implicit solver are needed for each time step.
\subsubsection{PD-ARS with SSPRK3}
Here we give the optimized PD-ARS with SSPRK3.
Its standard double Butcher tableau are (explicit tableau on the left, implicit tableau on the right)
\begin{equation}
  \begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6 & 1/6 & 2/3\\\hline
  	  & 0 & 1/6 & 1/6 & 2/3
  \end{array}
\end{equation}
For this schemes, $c_{\mbox{\tiny Sch}}= 1$.
Three implicit solver are taken for each time step.
Since PD-ARS with SSPRK3 is as accurate as PD-ARS with SSPRK2 in collision involved region, see Section~\ref{se:NumericalTests} for the plots, it may not offer any advantage.

>>>>

Since the set of realizable moments is convex, convex-invariant schemes, which maintain states in a convex set, can be used to design realizability-preserving methods for the two-moment model. 
Ideally, the scheme should also be high-order accurate and work well in the asymptotic diffusion limit (characterized by frequent collisions and long time scales).  
The following discussion considers the construction of such convex-invariant schemes.

@@

\subsection{Convex-Invariant Implicit-Explicit Schemes}

@@

By requiring the IMEX scheme be GSA, $\vect{u}^{n+1} = \vect{u}^{(s)}$.
Then using those Lemma proved in \cite{chu_etal_2018}, we can find those constraints, such as Eq.~\eqref{eq:convexInvariant} for type~ARS, that $c_{ij}$ and $\hat{c}_{ij}$ (or $a_{ij}$ and $\tilde{a}_{ij}$) need to satisfy to ensure $\vect{u}^{(i)}$, Eq.~\eqref{eq:imexStagesRewrite}, be realizable.
Hence $\vect{u}^{(i)}$ and $\vect{u}^{n+1}$ are realizable and the method with this convex-invariant IMEX scheme is realizability-preserving.

@@

Unfortunately, no coefficients can be found for standard IMEX schemes under the convex-invariant constraints and the accuracy constraints for 2nd-order or higher. 
To circumvent this problem, some correction steps are introduced after the assembly step, Eq.~\eqref{imexIntermediate}, such as Hu et al.\cite{hu_etal_2018}.
However, the correction steps impose extra constraints for realizability on time step.
In the circumstance that we have, correction steps can ruin the efficiency we gain from the IMEX scheme.
To keep things simple, we aim for 1st-order in general, high-order in streaming limit, diffusion accurate, convex-invariant IMEX schemes.

\subsection{Diffusion Accurate, Convex-Invariant Implicit-Explicit Schemes}

Diffusion accurate is an important property the time integration should have.
In the diffusion limit, the distribution function is nearly isotropic so that $\vect{\cK}\approx\f{1}{3}\,\cJ\,\vect{I}$ and $\vect{\cH}\approx-\f{1}{3}\,\tau\,\nabla\cJ$.
The moment behavior are governed by (e.g., \cite{jinLevermore_1996})
\begin{equation}
  \pd{\cJ}{t} + \nabla\cdot\vect{\cH} = 0
  \quad\text{and}\quad
  \vect{\cH} = - \tau\,\nabla\cdot\vect{\cK}.  
  \label{eq:diffusionLimit}
\end{equation}
In the context of IMEX schemes, the above relationships imply that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold. 
That's
\begin{equation}
   \vect{e}_{i}^{T}A^{-1}\tilde{A}\,\vect{e} = 1, \quad i=2,\ldots,s,
\end{equation}
where $\vect{e}_{i}$ is the $i$th column of the $s\times s$ identity matrix.
It can be rewritten as $c=\tilde{c}$.
We have proved in \cite{chu_etal_2018} that only type~ARS IMEX schemes can be diffusion accurate and convex-invariant at the same time.

@@

Combining the conditions we have, the 1st-order in general, high-order in streaming limit, diffusion accurate, convex-invariant IMEX schemes we need are type~ARS IMEX schemes satisfying the following constraints:
\begin{enumerate}
    \item Consistency of the implicit coefficients:
    \begin{equation}
      \sum_{i=1}^{s}w_{i}=1.
    \end{equation}
    \item High-order accuracy in the streaming limit, 2nd-order for example:
    \begin{equation}
      \sum_{i=1}^{s}\tilde{w}_{i}=1
      \quad\text{and}\quad
      \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2}.
      \label{eq:orderConditionsEx}
    \end{equation}
    \item Diffusion accurate:
    \begin{equation}
      c=\tilde{c}.
      \label{eq:diffusionCondition}
    \end{equation}
    \item Convex-invariant:
    \begin{align}
      &a_{ii}>0, \quad c_{i0},\tilde{c}_{i0}\ge0, \quad \text{for} \quad i=2,\ldots,s, \nonumber \\
      &\text{and} \quad c_{ij},\tilde{c}_{ij}\ge0, \quad \text{for} \quad i=3,\ldots,s, \quad\text{and}\quad j=2,\ldots,i-1.  
      \label{eq:convexInvariant}
    \end{align}
    with $\sum_{j=0}^{i-1}c_{ij}=1$, for $i=1,\ldots,s$, and $c_{\Sch}:=\min_{\substack{i = 2,\ldots,s \\ 
                  j = 0,2,\ldots,i-1}}\,\f{1}{\hat{c}_{ij}}>0$.
    \item Less than five stages ($s\le4$) \label{cod:statges}.
    \item Globally stiffly accurate: $a_{si}=w_{i}$ and $\tilde{a}_{si}=\tilde{w}_{i},\quad i=1,\ldots,s$. 
\end{enumerate}  
The \ref{cod:statges} constraint is given by efficiency consideration.
We call the IMEX scheme satisfying the above conditions {PD-ARS}. (Definition 3 in \cite{chu_etal_2018}.)
We give two optimized PD-ARS schemes here: PD-ARS with SSPRK2 and  PD-ARS with SSPRK3. 
They have second- and third-order accuracy in the streaming limit, respectively.
\subsubsection{PD-ARS with SSPRK2}
Here we give the optimized PD-ARS with SSPRK2 in the standard double Butcher tableau form: explicit tableau on the left ($\tilde{A}$), implicit tableau on the right ($A$):
\begin{equation}
  \begin{array}{c | c c c}
  	0 & 0   & 0 & 0 \\
  	1 & 1   & 0 & 0 \\
  	1 & 1/2 & 1/2 & 0 \\ \hline
  	  & 1/2 & 1/2 & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1 & 0 & 1/2 & 1/2 \\ \hline
  	  & 0 & 1/2 & 1/2
  \end{array}
\end{equation}
For this scheme, $c_{\mbox{\tiny Sch}}= 1$ and only two implicit solver are needed for each time step.
\subsubsection{PD-ARS with SSPRK3}
Here we give the optimized PD-ARS with SSPRK3.
Its standard double Butcher tableau are (explicit tableau on the left, implicit tableau on the right)
\begin{equation}
  \begin{array}{c | c c c c}
  	    &     &     &     &  \\
  	 1  & 1   &     &     &  \\
  	1/2 & 1/4 & 1/4 &  \\
  	 1  & 1/6 & 1/6 & 2/3 &  \\ \hline
  	    & 1/6 & 1/6 & 2/3 &
  \end{array}
  \qquad
  \begin{array}{c | c c c c}
  	0 & 0 & 0            & 0            \\
  	1 & 0 & 1            & 0            \\
  	1/2 & 0 & 1/4 & 1/4 \\ 
  	1 & 0 & 1/6 & 1/6 & 2/3\\\hline
  	  & 0 & 1/6 & 1/6 & 2/3
  \end{array}
\end{equation}
For this schemes, $c_{\mbox{\tiny Sch}}= 1$.
Three implicit solver are taken for each time step.
Since PD-ARS with SSPRK3 is as accurate as PD-ARS with SSPRK2 in collision involved region, see Section~\ref{se:NumericalTests} for the plots, it may not offer any advantage.

### 2018-11-07 Documents/M1/Astronum_2018/sections/TwoMoments.tex
Changed constraint-preserving to realizability-preserving.

The constraints in Eq.~\eqref{eq:MomentsBounds} define realizable moments $\bcM$.  
For fermions, realizable moments can only be constructed from a distribution satisfying the bounds $f \in [0,1]$.  
Moreover, the set of realizable moments is convex.  
As we will see later in Section~\ref{se:SpacialDiscretization}, this convexity makes it possible to design a constraint-preserving discretization for solving the two-moment model numerically.

>>>>

The constraints in Eq.~\eqref{eq:MomentsBounds} define realizable moments $\bcM$.  
For fermions, realizable moments can only be constructed from a distribution satisfying the bounds $f \in [0,1]$.  
Moreover, the set of realizable moments is convex.  
As we will see later in Section~\ref{se:SpacialDiscretization}, this convexity makes it possible to design a realizability-preserving discretization for solving the two-moment model numerically.

### 2018-11-07 Documents/M1/Astronum_2018/sections/intro.tex
Changed constraint-preserving to realizability-preserving.

Applying the two-moment method alone does not guarantee an affordable solution.
To be precise, neutrinos interact with the background, and the characteristic time scale of these interactions, which can be $\sim\mathcal{O}(10^{-13})$~second, is short compared to the duration of the CCSN explosion, which $\sim\mathcal{O}(1)$~second.  
This makes the time step needed by an explicit time integrator prohibitively small. 
On the other hand, solving the moment equations fully implicitly requires inverting globally many band-structured matrices whose sizes depend on the spatial discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require fewer time steps comparing with fully explicit method, and the calculation for each step is easily parallelizable.  
For a non-relativistic system where the propagating speed of gravity wave and fluid is much slower than the light speed, fully implicit method is a better choice for neutrino transport in CCSN simulation.\cite{cernohorskyVanWeert_1992}(read ?)
But for a relativistic circumstance that we have, where the gravity, fluid, and neutrino have a relativistic propagating speed, IMEX method
balances the efficiency and cost.

To model neutrino transport using two-moment method, two things need to be chosen carefully: an algebraic closure based on Fermi-Dirac statistic for closing the two-moment equations and a convex-invariant, diffusion-accurate IMEX scheme to ensure a physical result.
Since the neutrino distribution function is bounded ($f\in[0,1]$) by the Pauli exclusion principle, its moments as weight integrals of a bounded function over domain $\omega\in\bbS^{2}$ are also bounded.
We call the moments satisfying the constraints due to Fermi-Dirac statistics \textit{realizable moments}.
The algebraic closure should give a realizable $\bcK$, and the well-posedness of the closure requires realizable $\cJ$ and $\bcH$.
It explains why an algebraic closure based on Fermi-Dirac statistics is needed.
Realizability of $\cJ$ and $\bcH$ after each time step requires a convex-invariant IMEX scheme.
Since the realizable moments form a convex set, it is possible to construct a constraint-preserving IMEX scheme for two-moment neutrino transport.
What's more, the physics of neutrino transport in CCSN requires the IMEX scheme to be diffusion-accurate.

The study of moment realizability and constraint-preserving, diffusion-accurate IMEX schemes motivate this work.
Gottlieb et al\cite{gottlieb_etal_2001} showed standard IMEX scheme with strong-stability-preserving can't have an order higher than first without a restrict time step requirement.
To have second-order (or higher-order) accuracy, some correction steps are needed.
Unfortunately, the correction steps can deteriorate the accuracy of the IMEX scheme in the diffusion limit or restrict the time step.
To keep things simple, we focus on IMEX schemes without correction steps and require them to be high-order (second or higher order) in the streaming limit and diffusion-accurate.
We call these IMEX schemes \textit{PD-ARS}.

@@

\rc{Reference check}

>>>>

Applying the two-moment method alone does not guarantee an affordable solution.
To be precise, neutrinos interact with the background, and the characteristic time scale of these interactions, which can be $\sim\mathcal{O}(10^{-13})$~second, is short compared to the duration of the CCSN explosion, which $\sim\mathcal{O}(1)$~second.  
This makes the time step needed by an explicit time integrator prohibitively small. 
On the other hand, solving the moment equations fully implicitly requires inverting globally many band-structured matrices whose sizes depend on the spatial discretization.
Such a global inversion is both expensive and unfriendly to parallelization.
To circumvent these challenges, implicit-explicit (IMEX) methods are taken into consideration.
By treating the transport terms in the two-moment equations explicitly and the collision terms implicitly, IMEX methods are subject only to a time step governed by the explicit transport terms, and the matrices to be inverted are block diagonal.
Therefore, IMEX methods require fewer time steps comparing with fully explicit method, and the calculation for each step is easily parallelizable.  
%For a non-relativistic system where the propagating speed of gravitional wave and fluid is much slower than the light speed, fully implicit method is a better choice for neutrino transport in CCSN simulation.\cite{cernohorskyVanWeert_1992}(read ?)
For a relativistic circumstance that we have, where the gravity, fluid, and neutrino have a relativistic propagating speed, IMEX method is an efficient method.

To model neutrino transport using two-moment method, two things need to be chosen carefully: an algebraic closure based on Fermi-Dirac statistic for closing the two-moment equations and a convex-invariant, diffusion-accurate IMEX scheme to ensure a physical result.
Since the neutrino distribution function is bounded ($f\in[0,1]$) by the Pauli exclusion principle, its moments as weight integrals of a bounded function over domain $\omega\in\bbS^{2}$ are also bounded.
We call the moments satisfying the constraints due to Fermi-Dirac statistics \textit{realizable moments}.
The algebraic closure should give a realizable $\bcK$, and the well-posedness of the closure requires realizable $\cJ$ and $\bcH$.
It explains why an algebraic closure based on Fermi-Dirac statistics is needed.
Realizability of $\cJ$ and $\bcH$ after each time step requires a convex-invariant IMEX scheme.
Since the realizable moments form a convex set, it is possible to construct a realizability-preserving method with convex-invariant IMEX scheme for two-moment neutrino transport.
What's more, the physics of neutrino transport in CCSN requires the IMEX scheme to be diffusion-accurate.

The study of moment realizability and realizability-preserving method with diffusion-accurate IMEX schemes motivate this work.
Gottlieb et al\cite{gottlieb_etal_2001} showed standard IMEX scheme with strong-stability-preserving can't have an order higher than first without a restrict time step requirement.
To have second-order (or higher-order) accuracy, some correction steps are needed.
Unfortunately, the correction steps can deteriorate the accuracy of the IMEX scheme in the diffusion limit or restrict the time step.
To keep things simple, we focus on IMEX schemes without correction steps and require them to be high-order (second or higher order) in the streaming limit and diffusion-accurate.
We call these IMEX schemes \textit{PD-ARS}.

@@

%\rc{Reference check}
